---
title: The three tables in genomics
author: Jeff Leek
output:
  rmarkdown::html_document:
    toc: true
  vignette: >  
    %\VignetteIndexEntry{The three tables in genomics}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r front, child="./../front.Rmd", echo=FALSE}
```


## Dependencies

This document depends on the following packages:

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
   library(devtools)
  library(Biobase)
})
```

```{r load}
  library(devtools)
  library(Biobase)
```


To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r, eval=FALSE}
install.packages("devtools")
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("Biobase"))
```


## Load the data from the web

```{r load_data}
con=url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
```

## An expression set

This kind of object in R contains the three tables

```{r expression_set}
bm = bodymap.eset
bm
```

### Table 1: the genomics data

These are usually high dimensional measurements. In this case it is RNA-sequencing data. Each row is a gene and each column is a sample. 
```{r expression_data}
exp_data = exprs(bm)
dim(exp_data)
head(exp_data,n=5)
```

### Table 2: the phenotype data

This is the "meta data" or "phenotype data" that is what you typically associate the genomic data with. Each column is a variable and each row is a sample.

```{r pheno_data}
pheno_data = pData(bm)
dim(pheno_data)
head(pheno_data)
```

### Table 3: the feature data

This is information about the genes or other genomic features you are measuring. Each column is a variable and each row is a gene/feature. 

```{r feature_data}
feature_data = fData(bm)
dim(fData(bodymap.eset))
fData(bodymap.eset)[1:10,,1]
```

## Session information

Here is the session information 

```{r session_info}
devtools::session_info()
```

It is also useful to compile the time the document was processed. This document was processed on: `r Sys.Date()`.



