---
title: Data transforms
author: Jeff Leek
output:
  rmarkdown::html_document:
    toc: true
  vignette: >  
    %\VignetteIndexEntry{Data transforms}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r front, child="./../front.Rmd", echo=FALSE}
```


## Dependencies

This document depends on the following packages:

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
  library(dplyr)
})
```

```{r load}
  library(devtools)
  library(Biobase)
  library(dplyr)
```


To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r install_packages, eval=FALSE}
install.packages(c("devtools"))
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("Biobase"))
```


## General principles

* Make sure data is on the right scale for
  * Visualization
  * Statistical modeling


## Load some data

We will use this expression set to look at how we use plots and tables to check for different characteristics

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata=pData(bm)
edata=as.data.frame(exprs(bm))
fdata = fData(bm)
ls()
```

## Skewed distributions

We would like continuous data to be nice and symmetric like a normal distribution for two reasons:(1) plots are easier to see this way and (2) most statistical methods are designed to work better for non-skewed data. 

```{r}
hist(rnorm(1000),col=2)
```

Realistically though most genomic data is skewed

```{r}
hist(edata[,1],col=2,breaks=100)
```

One way to address this skew is to use a transformation. One common transformation is the `log` transform. 

```{r}
hist(log(edata[,1]),col=2,breaks=100)
```

One thing to be careful with is that values of zero become `-inf` when you apply the log transform because the log of zero isn't defined. 

```{r}
min(log(edata))
```

You can remove this problem by adding a small number to all the counts before taking the log. This doesn't change the overall distribution much but resolves the zero problem

```{r}
min(log(edata[,1] + 1))
hist(log(edata[,1] + 1),breaks=100,col=2)
```

A common choice is to use the `log2` transform. The reason is that data on the `log2` scale can be expressed as [fold changes](https://en.wikipedia.org/wiki/Fold_change). If `log2(x) - log2(y) = z` then there is a `z` fold change difference between `x` and `y`. 

```{r}
hist(log2(edata[,1] + 1),breaks=100,col=2)
```


Let's zoom in on the values greater than zero (we'll come back to the zero values in a minute)

```{r}
hist(log2(edata[,1] + 1),breaks=100,col=2,xlim=c(1,15),ylim=c(0,400))
```

This is starting to look a little better but we still have the zeros problem. 


## Removing features without much data

In many types of sequencing data we observe that there is not much data for many features. 

```{r}
hist(rowSums(edata==0),col=2)
```

A common pre-processing technique is to remove features that don't have much data. One way to identify and remove them is to use `rowMeans` and `rowMedians`

```{r}
low_genes = rowMeans(edata) < 5
table(low_genes)
filt_edata = filter(edata,!low_genes)
dim(filt_edata)

low_genes2 = rowMedians(as.matrix(edata)) < 5
table(low_genes2,low_genes)
filt_edata2 = filter(edata,!low_genes2)
dim(filt_edata2)
```


After filtering we observe that using the `log` transform makes the distribution much more symmetric but there are still zeros. We could increase the filter to remove this issue if we needed to. 

```{r}
hist(log2(filt_edata[,1] + 1),col=2)
```


## More transforms

There are a large number of other transforms depending on the type of data you are using. Some common ones are:

* [Variance stabilizing transforms](https://en.wikipedia.org/wiki/Variance-stabilizing_transformation) which seek to remove a mean variance relationship among the data
* [Box-Cox transforms](https://en.wikipedia.org/wiki/Power_transform) which seek to make the data approximately Normally distributed
* [rlog transform](http://rpackages.ianhowson.com/bioc/DESeq2/man/rlog.html) - unique to genomics count data, this is a regularized version of the log transform that seeks to minimize differences at low count levels. 


## Session information

Here is the session information 

```{r session_info}
devtools::session_info()
```

It is also useful to compile the time the document was processed. This document was processed on: `r Sys.Date()`.
