---
title: Galton example
author: Jeff Leek
output:
  rmarkdown::html_document:
    toc: true
  vignette: >  
    %\VignetteIndexEntry{Galton example}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r front, child="./../front.Rmd", echo=FALSE}
```


## Dependencies

This document depends on the following packages:

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
  library(UsingR)
})
```

```{r load}
  library(devtools)
  library(Biobase)
  library(UsingR)
```


To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r install_packages, eval=FALSE}
install.packages(c("devtools","UsingR"))
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("Biobase"))
```


### Load Galton Data

```{r galton,fig.height=4,fig.width=8}
data(galton)
par(mfrow=c(1,2))
hist(galton$child,col="blue",breaks=100)
hist(galton$parent,col="blue",breaks=100)
```

### The distribution of child heights

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
```

### Only know the child  - average height

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
meanChild <- mean(galton$child)
lines(rep(meanChild,100),seq(0,150,length=100),col="red",lwd=5)
```


### What if we plot child versus average parent

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
```

### Average parent = 65 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near65 <- galton[abs(galton$parent - 65)<1, ]
points(near65$parent,near65$child,pch=19,col="red")
lines(seq(64,66,length=100),rep(mean(near65$child),100),col="red",lwd=4)
```

### Average parent = 71 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near71 <- galton[abs(galton$parent - 71)<1, ]
points(near71$parent,near71$child,pch=19,col="red")
lines(seq(70,72,length=100),rep(mean(near71$child),100),col="red",lwd=4)
```


### Fitting a line



```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lm1 <- lm(galton$child ~ galton$parent)
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```


### Not all points are on the line

```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

### Plot what is leftover

```{r, dependson="lm1",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
plot(galton$parent,lm1$residuals,col="blue",pch=19)
abline(c(0,0),col="red",lwd=3)
```



## Session information

Here is the session information 

```{r session_info}
devtools::session_info()
```

It is also useful to compile the time the document was processed. This document was processed on: `r Sys.Date()`.



