---
output:
  BiocStyle::html_document
---


<!-- These options are set up specifically for Bioconductor flavored Markdown --> 


<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, results = 'asis',include=FALSE}
BiocStyle::markdown()
```

```{r global_options,include=FALSE}
## see ch. 10 Hooks of Xie's knitr book
knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 19)
})
opts_chunk$set(setPch = TRUE)
library(RSkittleBrewer)
# Make the colors pretty
trop = RSkittleBrewer("tropical")
palette(trop)
```


# Poverty example

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`

### WHO childhood hunger data


```{r whoDataLoad}
hunger = read.csv("http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*")
hunger <- hunger[hunger$Sex!="Both sexes",]
head(hunger)
```

### Plot percent hungry versus time

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
```


### Add the linear model

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
lines(hunger$Year,lm1$fitted,lwd=3,col="darkgrey")
```


### Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
```

### Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmM <- lm(hunger$Numeric[hunger$Sex=="Male"] ~ hunger$Year[hunger$Sex=="Male"])
lmF <- lm(hunger$Numeric[hunger$Sex=="Female"] ~ hunger$Year[hunger$Sex=="Female"])
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
lines(hunger$Year[hunger$Sex=="Male"],lmM$fitted,col="black",lwd=3)
lines(hunger$Year[hunger$Sex=="Female"],lmF$fitted,col="red",lwd=3)
```

### Two lines, same slope in R


```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] ),col="black",lwd=3)
```


### Two lines, different slopes in R


```{r lmBothChunk, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex + hunger$Sex*hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] +lmBoth$coeff[4]),col="black",lwd=3)
```


### Two lines, different slopes in R


```{r, dependson="lmBothChunk",fig.height=4,fig.width=4}
summary(lmBoth)
```


### Dependencies

These are the packages that this tutorial depends on:

* [BiocStyle](http://www.bioconductor.org/packages/release/bioc/html/BiocStyle.html)
* [knitr](https://cran.r-project.org/package=knitr)
* [RSkittleBrewer](https://github.com/alyssafrazee/RSkittleBrewer)
* [Biobase](http://bioconductor.org/packages/release/bioc/html/Biobase.html)
* [devtools](https://cran.r-project.org/package=devtools)
* [dplyr](https://cran.r-project.org/package=dplyr)
* [org.Hs.eg.db](http://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html)
* [AnnotationDbi](http://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html)


### Session information

```{r session_info}
devtools::session_info()
```


