---
title: Many regressions
author: Jeff Leek
output:
  rmarkdown::html_document:
    toc: true
  vignette: >  
    %\VignetteIndexEntry{Many regressions}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r front, child="./../front.Rmd", echo=FALSE}
```


## Dependencies

This document depends on the following packages:

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
  library(limma)
  library(edge)
})
```

```{r load}
  library(devtools)
  library(Biobase)
  library(limma)
  library(edge)
```


To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r install_packages, eval=FALSE}
install.packages(c("devtools"))
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("Biobase","limma","jdstorey/edge"))
```


## Download the data


Here we are going to use some data from the paper [Detection of redundant fusion transcripts as biomarkers or disease-specific therapeutic targets in breast cancer.](http://www.ncbi.nlm.nih.gov/pubmed/22496456) that uses data from different normal human tissues (called the Illumina BodyMap data).

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
pdata=pData(bot)
edata=as.matrix(exprs(bot))
fdata = fData(bot)
ls()
```

## Transform the data

Here we will transform the data and remove lowly expressed genes. 

```{r}
edata = log2(as.matrix(edata) + 1)
edata = edata[rowMeans(edata) > 10, ]
```

## Fit many regression models at once. 

```{r}
mod = model.matrix(~ pdata$strain)
fit = lm.fit(mod,t(edata))
names(fit)
```

### Compare to output of `lm` 

```{r}
fit$coefficients[,1]
tidy(lm(as.numeric(edata[1, ]) ~ pdata$strain))
```


### Look at the coefficients across genes

```{r}
par(mfrow=c(1,2))
hist(fit$coefficients[1,],breaks=100,col=2,xlab="Intercept")
hist(fit$coefficients[2,],breaks=100,col=2,xlab="Strain")
abline(v=0,lwd=3,col=1)
```

### Look at the residuals for a couple of genes

```{r}
par(mfrow=c(1,2))
plot(fit$residuals[,1],col=2)
plot(fit$residuals[,2],col=2)
```


### Fit many regressions with an adjustment

```{r}
mod_adj = model.matrix(~ pdata$strain + as.factor(pdata$lane.number))
fit_adj = lm.fit(mod_adj,t(edata))
fit_adj$coefficients[,1]
```


## Fit many regressions with the limma package

```{r}
fit_limma = lmFit(edata,mod_adj)
names(fit_limma)
fit_limma$coefficients[1,]
fit_adj$coefficients[,1]
```

## Fit many regressions with the edge package

```{r}
edge_study = build_study(data=edata,grp=pdata$strain,adj.var=as.factor(pdata$lane.number))
fit_edge = fit_models(edge_study)
summary(fit_edge)
fit_edge@beta.coef[1,]
fit_limma$coefficients[1,]
```


## More information

You can find a lot more information on this model fitting strategy in:

* The [limma paper](http://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)
* The [limma vignette](http://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)

## Session information

Here is the session information 

```{r session_info}
devtools::session_info()
```

It is also useful to compile the time the document was processed. This document was processed on: `r Sys.Date()`.



