---
output:
  BiocStyle::html_document
---

<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
palette(RSkittleBrewer::RSkittleBrewer("tropical"))
```

# Representing Data in R

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`

### Objects and operations

* Vectors, Matrices, Data frames, Lists, Factors, Missing values
* Subsetting, Logical subsetting

### Characters

```{r}
firstName = "jeff"
class(firstName)
firstName
```

### Numeric
```{r}
heightCM = 188.2
class(heightCM)
heightCM
```


### Integer
```{r}
numberSons = 1L
class(numberSons)
numberSons
```

### Logical
```{r}
teachingCoursera = TRUE
class(teachingCoursera)
teachingCoursera
```

### Vectors
A set of values with the same class
```{r}
heights = c(188.2, 181.3, 193.4)
heights

firstNames = c("jeff","roger","andrew","brian")
firstNames

```

### Lists
A vector of values of possibly different classes
```{r}
vector1 = c(188.2, 181.3, 193.4)
vector2 = c("jeff","roger","andrew","brian")
myList = list(heights=vector1,firstNames=vector2)
myList

```

### Matrices
Vectors with multiple dimensions
```{r}
myMatrix = matrix(c(1,2,3,4),byrow=T,nrow=2)
myMatrix

```

### Data frames
Multiple vectors of possibly different classes, of the same length

```{r}
vector1 = c(188.2,181.3,193.4,192.3)
vector2 = c("jeff","roger","andrew","brian")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)
myDataFrame
```

### Factors
Qualitative variables that can be included in models

```{r}
smoker = c("yes","no","yes","yes")
smokerFactor = as.factor(smoker)
smokerFactor

```

### Missing values

In R they are usually coded NA
```{r}
vector1 = c(188.2,181.3,193.4,NA)
vector1
is.na(vector1)
```

### Subsetting

```{r}
vector1 = c(188.2,181.3,193.4,192.3)
vector2 = c("jeff","roger","andrew","brian")
myDataFrame = data.frame(heights=vector1,firstNames=vector2)

vector1[1]
vector1[c(1,2,4)]
```

### Subsetting

```{r}
myDataFrame[1,1:2]
myDataFrame$firstNames

```

### Logical subsetting
```{r}
myDataFrame[myDataFrame$firstNames=="jeff",]
myDataFrame[myDataFrame$heights < 190,]
```

### Variable naming conventions

Variable names should be short, but descriptive. Here are some common styles

__Camel caps__
```{r}
myHeightCM = 188
```
__Underscore__
```{r}
my_height_cm = 188
```
__Dot separated__
```{r}
my.height.cm = 188
```


### Style guides

* [http://4dpiecharts.com/r-code-style-guide/](http://4dpiecharts.com/r-code-style-guide/)
* [http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html)
* [http://wiki.fhcrc.org/bioc/Coding_Standards](http://wiki.fhcrc.org/bioc/Coding_Standards)
* For more information see this lecture on data types [Data Types](http://www.youtube.com/watch?v=5AQM-yUX9zg&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=5)
