---
output:
  BiocStyle::html_document
---


<!-- These options are set up specifically for Bioconductor flavored Markdown --> 


<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, results = 'asis',include=FALSE}
BiocStyle::markdown()
```

```{r global_options,include=FALSE}
## see ch. 10 Hooks of Xie's knitr book
knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 19)
})
opts_chunk$set(setPch = TRUE)
library(RSkittleBrewer)
# Make the colors pretty
trop = RSkittleBrewer("tropical")
palette(trop)
```


# Linear models

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`


### General principles

* Fitting "lines" to samples
* Only linear relationships are captured
* Be careful about outlier points
* Be careful about distributions of residuals 




### Dependencies

These are the packages that this tutorial depends on:

* [BiocStyle](http://www.bioconductor.org/packages/release/bioc/html/BiocStyle.html)
* [knitr](https://cran.r-project.org/package=knitr)
* [RSkittleBrewer](https://github.com/alyssafrazee/RSkittleBrewer)
* [Biobase](http://bioconductor.org/packages/release/bioc/html/Biobase.html)
* [devtools](https://cran.r-project.org/package=devtools)
* [dplyr](https://cran.r-project.org/package=dplyr)
* [org.Hs.eg.db](http://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html)
* [AnnotationDbi](http://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html)


### Session information

```{r session_info}
devtools::session_info()
```


