---
output:
  BiocStyle::html_document
---


<!-- These options are set up specifically for Bioconductor flavored Markdown --> 


<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, results = 'asis',include=FALSE}
BiocStyle::markdown()
```

```{r global_options,include=FALSE}
## see ch. 10 Hooks of Xie's knitr book
knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 19)
})
opts_chunk$set(setPch = TRUE)
library(RSkittleBrewer)
# Make the colors pretty
trop = RSkittleBrewer("tropical")
palette(trop)
```


# Galton example

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`

### Load Galton Data

```{r galton,fig.height=4,fig.width=8}
library(UsingR); data(galton)
par(mfrow=c(1,2))
hist(galton$child,col="blue",breaks=100)
hist(galton$parent,col="blue",breaks=100)
```

### The distribution of child heights

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
```

### Only know the child  - average height

```{r, dependson="galton",fig.height=4,fig.width=4}
hist(galton$child,col="blue",breaks=100)
meanChild <- mean(galton$child)
lines(rep(meanChild,100),seq(0,150,length=100),col="red",lwd=5)
```


### What if we plot child versus average parent

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
```

### Average parent = 65 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near65 <- galton[abs(galton$parent - 65)<1, ]
points(near65$parent,near65$child,pch=19,col="red")
lines(seq(64,66,length=100),rep(mean(near65$child),100),col="red",lwd=4)
```

### Average parent = 71 inches tall

```{r, dependson="galton",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
near71 <- galton[abs(galton$parent - 71)<1, ]
points(near71$parent,near71$child,pch=19,col="red")
lines(seq(70,72,length=100),rep(mean(near71$child),100),col="red",lwd=4)
```


### Fitting a line



```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lm1 <- lm(galton$child ~ galton$parent)
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```


### Not all points are on the line

```{r, dependson="lm1",fig.height=4,fig.width=4}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

### Plot what is leftover

```{r, dependson="lm1",fig.height=4,fig.width=8}
par(mfrow=c(1,2))
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
plot(galton$parent,lm1$residuals,col="blue",pch=19)
abline(c(0,0),col="red",lwd=3)
```


### Dependencies

These are the packages that this tutorial depends on:

* [BiocStyle](http://www.bioconductor.org/packages/release/bioc/html/BiocStyle.html)
* [knitr](https://cran.r-project.org/package=knitr)
* [RSkittleBrewer](https://github.com/alyssafrazee/RSkittleBrewer)
* [Biobase](http://bioconductor.org/packages/release/bioc/html/Biobase.html)
* [devtools](https://cran.r-project.org/package=devtools)
* [dplyr](https://cran.r-project.org/package=dplyr)
* [org.Hs.eg.db](http://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html)
* [AnnotationDbi](http://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html)


### Session information

```{r session_info}
devtools::session_info()
```


