---
title: Galton example
author: Jeff Leek
output:
  rmarkdown::html_document:
    toc: true
  vignette: >  
    %\VignetteIndexEntry{Galton example}
    %\VignetteEngine{knitr::rmarkdown}
    \usepackage[utf8]{inputenc}
---

```{r front, child="./../front.Rmd", echo=FALSE}
```


## Dependencies

This document depends on the following packages:

```{r load_hidden, echo=FALSE, results="hide", warning=FALSE}
suppressPackageStartupMessages({
  library(devtools)
  library(Biobase)
})
```

```{r load}
  library(devtools)
  library(Biobase)
```


To install these packages you can use the code (or if you are compiling the document, remove the `eval=FALSE` from the chunk.)

```{r install_packages, eval=FALSE}
install.packages(c("devtools"))
source("http://www.bioconductor.org/biocLite.R")
biocLite(c("Biobase"))
```

## WHO childhood hunger data


```{r whoDataLoad}
hunger = read.csv("http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*")
hunger <- hunger[hunger$Sex!="Both sexes",]
head(hunger)
```

## Plot percent hungry versus time

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
```


## Add the linear model

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
lines(hunger$Year,lm1$fitted,lwd=3,col="darkgrey")
```


## Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
```

## Color by male/female

```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmM <- lm(hunger$Numeric[hunger$Sex=="Male"] ~ hunger$Year[hunger$Sex=="Male"])
lmF <- lm(hunger$Numeric[hunger$Sex=="Female"] ~ hunger$Year[hunger$Sex=="Female"])
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
lines(hunger$Year[hunger$Sex=="Male"],lmM$fitted,col="black",lwd=3)
lines(hunger$Year[hunger$Sex=="Female"],lmF$fitted,col="red",lwd=3)
```

## Two lines, same slope in R


```{r, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] ),col="black",lwd=3)
```


## Two lines, different slopes in R


```{r lmBothChunk, dependson="whoDataLoad",fig.height=4,fig.width=4}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex + hunger$Sex*hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] +lmBoth$coeff[4]),col="black",lwd=3)
```


## Two lines, different slopes in R


```{r, dependson="lmBothChunk",fig.height=4,fig.width=4}
summary(lmBoth)
```



## Session information

Here is the session information 

```{r session_info}
devtools::session_info()
```

It is also useful to compile the time the document was processed. This document was processed on: `r Sys.Date()`.



