---
output:
  BiocStyle::html_document
---


<!-- These options are set up specifically for Bioconductor flavored Markdown --> 


<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, results = 'asis',include=FALSE}
BiocStyle::markdown()
```

```{r global_options,include=FALSE}
## see ch. 10 Hooks of Xie's knitr book
knit_hooks$set(setPch = function(before, options, envir) {
  if(before) par(pch = 19)
})
opts_chunk$set(setPch = TRUE)
library(RSkittleBrewer)
# Make the colors pretty
trop = RSkittleBrewer("tropical")
palette(trop)
```


# Many regressions

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`

### Download the data


Here we are going to use some data from the paper [Detection of redundant fusion transcripts as biomarkers or disease-specific therapeutic targets in breast cancer.](http://www.ncbi.nlm.nih.gov/pubmed/22496456) that uses data from different normal human tissues (called the Illumina BodyMap data).

```{r}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
bm = bodymap.eset
pdata=pData(bm)
edata=as.data.frame(exprs(bm))
fdata = fData(bm)
ls()
```

### Transform the data

Here we will transform the data and remove lowly expressed genes. 

```{r}
edata = log2(as.matrix(edata) + 1)
edata = edata[rowMeans(edata) > 10, ]
```

### Fit many regression models at once. 

```{r}
library(edge)
mod = model.matrix(~ pdata$num.tech.reps)
fit = lm.fit(mod,t(edata))
names(fit)
```

### Compare to output of `lm` 

```{r}
fit$coefficients[,1]
tidy(lm(edata[1, ] ~ pdata$num.tech.reps))
```


### Look at the coefficients across genes

```{r}
par(mfrow=c(1,2))
hist(fit$coefficients[1,],breaks=100,col=2,xlab="Intercept")
hist(fit$coefficients[2,],breaks=100,col=2,xlab="Tech Reps")
abline(v=0,lwd=3,col=1)
```

### Look at the residuals for a couple of genes

```{r}
par(mfrow=c(1,2))
plot(fit$residuals[,1],col=2)
plot(fit$residuals[,2],col=2)
```


### Dependencies

These are the packages that this tutorial depends on:

* [BiocStyle](http://www.bioconductor.org/packages/release/bioc/html/BiocStyle.html)
* [knitr](https://cran.r-project.org/package=knitr)
* [RSkittleBrewer](https://github.com/alyssafrazee/RSkittleBrewer)
* [Biobase](http://bioconductor.org/packages/release/bioc/html/Biobase.html)
* [devtools](https://cran.r-project.org/package=devtools)
* [dplyr](https://cran.r-project.org/package=dplyr)
* [org.Hs.eg.db](http://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html)
* [AnnotationDbi](http://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html)


### Session information

```{r session_info}
devtools::session_info()
```


