---
output:
  BiocStyle::html_document
---

<!-- rmarkdown v1 -->

<!--
%\VignetteIndexEntry{R Markdown Lecture}
%\VignettePackage{BiocStyle}
%\VignetteEngine{knitr::knitr}
-->

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
palette(RSkittleBrewer::RSkittleBrewer("tropical"))
```

# The three tables

Package: [genstats](https://github.com/jtleek/genstats)<br />
Author: [Jeff Leek](www.jtleek.com) <br />
Compilation date: `r Sys.Date()`


### Download the data

```{r}
con=url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)
```

### An expression set

This kind of object in R contains the three tables

```{r}
bm = bodymap.eset
bm
```

### Table 1: the genomics data

These are usually high dimensional measurements. In this case it is RNA-sequencing data. Each row is a gene and each column is a sample. 
```{r}
exp_data = exprs(bm)
dim(exp_data)
head(exp_data,n=5)
```

### Table 2: the phenotype data

This is the "meta data" or "phenotype data" that is what you typically associate the genomic data with. Each column is a variable and each row is a sample.

```{r}
pheno_data = pData(bm)
dim(pheno_data)
head(pheno_data)
```

### Table 3: the feature data

This is information about the genes or other genomic features you are measuring. Each column is a variable and each row is a gene/feature. 

```{r}
feature_data = fData(bm)
dim(fData(bodymap.eset))
fData(bodymap.eset)[1:10,,1]
```


### Session information

```{r session_info}
sessionInfo()
devtools::session_info()
```




